---
title: "easyNRD Performance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{easyNRD Performance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

The AHRQ Nationwide Readmissions Database (NRD) is an out-of-core workload by
default. A single year is roughly 15-18 million discharge rows, and each row is
wide: a typical analytical table includes core linkage columns, 40 diagnosis
columns, 25 procedure columns, weights, strata, and user-defined phenotype
flags.

That combination creates three practical performance constraints:

- the data are too large to load safely into R memory as data frames
- readmission linkage is a patient-year self-join over a very wide table
- ranking and tie-breaking introduce blocking SQL operators that can create
  large intermediates

`easyNRD` is designed around those constraints. The public pipeline stays lazy,
pushes work into DuckDB, and uses explicit internal checkpoints only where they
improve out-of-core stability.

# Pipeline Bottleneck

Across the current benchmark harness, `nrd_link_readmissions()` is the dominant
pipeline stage. That is expected.

Why it dominates:

- it performs a self-join within `YEAR` and `NRD_VISITLINK`
- it computes the HCUP gap formula for every candidate readmission pair
- it applies ordered tie-breaking over the surviving candidate set
- each ranking pass is a blocking operation from DuckDB's perspective

By contrast, `nrd_prepare()` is a simple arithmetic mutate and
`nrd_flag_condition()` stays cheap because predicate pushdown is already in good
shape.

`nrd_select()` and `nrd_export()` often appear expensive in wall-clock timings,
but those timings mostly reflect deferred upstream execution. They are usually
the point where DuckDB is finally forced to run the heavy linkage plan.

# Checkpoint Strategy

`easyNRD` now uses three deliberate internal `dplyr::compute()` checkpoints in
the readmission path:

1. a narrowed linkage checkpoint before the self-join
2. a checkpoint after the `gap` ranking pass
3. a checkpoint after the `DTE_cand` ranking pass

The final `KEY_NRD_cand` tie-break remains lazy and returns the final map
without another checkpoint.

This is intentional. On small synthetic fixtures, extra checkpointing can add a
little overhead. On real NRD-scale jobs, the benefit is that multiple blocking
window operators are no longer stacked into a single large DuckDB plan. That is
the key trade: slightly more small-data overhead in exchange for much better OOM
safety when the candidate map becomes large.

# Hardware Tuning

`easyNRD` keeps tuning internal to connection setup. No public function gains
extra arguments; all controls are environment-variable or option driven.

## Threads

Use either:

- `EASYNRD_MAX_THREADS`
- `options(easynrd.max_threads = ...)`

Environment variables take precedence over options. Values must be positive
integers. If you over-specify, `easyNRD` silently caps the request to the number
of physical cores detected by `parallel::detectCores(logical = FALSE)`.

```r
Sys.setenv(EASYNRD_MAX_THREADS = "8")

options(easynrd.max_threads = 8L)
```

## Memory Limit

Use either:

- `EASYNRD_MEMORY_LIMIT`
- `options(easynrd.memory_limit = ...)`

The value is passed through to DuckDB as `SET memory_limit = 'X'`.

```r
Sys.setenv(EASYNRD_MEMORY_LIMIT = "16GB")

options(easynrd.memory_limit = "16GB")
```

## Verbose Checkpoint Logging

Set `EASYNRD_VERBOSE=1` to emit checkpoint timing messages during linkage.

```r
Sys.setenv(EASYNRD_VERBOSE = "1")
```

When enabled, each internal checkpoint reports:

- the checkpoint label
- the row count after materialization
- elapsed seconds

This is useful while tuning hardware settings or diagnosing whether a job is
spending time in the narrowed base checkpoint versus the sequential ranking
passes.

# Parquet File Preparation

For analysts working with real NRD extracts, the most important performance work
happens before the first `easyNRD` call.

- stay parquet-native; never load NRD CSVs into R memory first
- keep files year-scoped, because `NRD_VISITLINK` is valid only within year
- ensure schema consistency across files for the same year
- ensure adequate temp-disk space, ideally at least 2-3x the size of the
  working dataset
- treat `nrd_export()` output as the canonical downstream artifact for later
  analysis, modeling, and survey work

These practices help DuckDB stream data efficiently and reduce avoidable temp
spill or schema-reconciliation overhead.

# Benchmark Harness

The developer benchmark harness lives at
`data-raw/benchmarks/benchmark_pipeline.R`.

It measures:

- fixture generation at two scales: 100K rows and 1M rows
- each major lazy pipeline stage independently
- the full end-to-end pipeline
- wall-clock time, `bench::mark()` timing, memory allocation, and basic temp-disk
  spill signals
- a `profvis` profile of `nrd_link_readmissions()` on the medium fixture

Run it from the repository root:

```{r benchmark-command, eval = FALSE}
Rscript data-raw/benchmarks/benchmark_pipeline.R
```

Sample console summary:

```text
Session summary:
- R version: R version 4.5.3 (2026-03-11 ucrt)
- DuckDB package: 1.4.1
- Physical cores: 8
- Available RAM: 25.1 GiB

Fixture summary:
# A tibble: 2 x 5
  scale   rows patients mean_stays_per_patient max_stays_per_patient
  medium 1e+05    34005                   2.94                     8
  large  1e+06   337706                   2.96                     8
```

The benchmark script is intentionally not executed during vignette builds.
Chunks that run the benchmark are marked `eval = FALSE`, and any code that does
run during render is lightweight.

# Practical Reading of Results

When you review benchmark output, focus on these questions:

- does `nrd_link_readmissions()` still dominate the pipeline?
- did a change reduce temp spill or shrink intermediate row counts?
- did a hardware setting materially improve the linkage stage?
- did a change only help small synthetic data, or does it actually reduce
  large-scale risk?

For `easyNRD`, the most important wins are often stability wins rather than
small-data speed wins. A slightly slower 100K fixture can still be the right
trade if it prevents memory blowups at 15-18 million rows.
